% English translation of: "Symbolic Dynamics — Exercises (2020), V. Berthé"
\documentclass[11pt,a4paper]{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{enumitem}
\usepackage{hyperref}

\title{Symbolic Dynamics\\Exercises}
\author{V.~Berth\'e \texttt{berthe@irif.fr}}
\date{2020}

\begin{document}
\maketitle

\section{Study of an infinite word defined by a substitution}

Let $A=\{a,b\}$ and let $\sigma : A \to A^\ast$ be the substitution defined by
\[
  \sigma(a)=abaa
  \qquad\text{and}\qquad
  \sigma(b)=babb.
\]

\begin{enumerate}[label=\arabic*.]
  \item How many fixed points does the substitution $\sigma$ admit in $A^{\mathbb{N}}$?
        In the rest, let $u$ denote the fixed point of $\sigma$ that begins with $a$.

  \item Show that the sequence $u$ is uniformly recurrent.

  \item Show that the set of its factors is invariant under exchanging the letters $a$ and $b$.

  \item List the factors of $u$ of length at most $6$.
        List the right special factors of $u$ of length $5$.
        Draw the Rauzy graphs of $u$ of order $1$ to $5$.
        What does the complexity of $u$ seem to be?

  \item Is the set of factors of $u$ invariant under reversal (mirror image)?

  \item The goal of this question is to prove the following decomposition lemma:
        every factor $W$ of $u$ of length at least $5$ can be written uniquely as
        \[
          W = R\,\sigma(V)\,S
        \]
        where $V$ is a factor of $u$,
        $R \in \{\varepsilon, a, aa, baa, b, bb, abb\}$, and
        $S \in \{\varepsilon, a, ab, aba, b, ba, bab\}$.
        \begin{enumerate}[label=(\alph*)]
          \item \textbf{Existence.}
                Show that if $u=u_0u_1\cdots u_n$, then
                \[
                  \sigma(u_n)=u_{4n}u_{4n+1}u_{4n+2}u_{4n+3}.
                \]
                Deduce the existence of such a decomposition.

          \item \textbf{Uniqueness.}
                Show that $R$ is determined by a prefix of $W$ of length at most $5$;
                to do this, by a case analysis, show that for each possible value of $R$,
                that value determines a prefix of $W$ of length at most $5$.
                Conclude by observing that determining $R$ implies determining $S$.
        \end{enumerate}

  \item Show that if $W$ is a right special factor of $u$ of length at least $5$,
        then $W$ decomposes uniquely as
        \[
          W = R\,\sigma(V),
        \]
        where $V$ is a right special factor of $u$ and $R \in \{\varepsilon, a, aa, baa, b, bb, abb\}$.
        You may use the fact that the suffix of $W$ of length $5$ is, in particular, right special.

  \item Show that for every $n\ge 2$, $u$ has the same number of right special factors of length $n$.
        Give this number.

  \item Deduce the complexity function of $u$.
\end{enumerate}

\section{Complexity functions of the form $p(n)=n+k$, for all $n$}

\begin{enumerate}[label=\arabic*.]
  \item Let $k\ge 1$. Construct, from the Fibonacci word, an example $u$ of a sequence whose
        complexity function $p_u(n)$ satisfies
        \[
          p_u(n)=n+k,\qquad \text{for all } n.
        \]

  \item Let $u$ be a sequence over the alphabet $A$ such that every letter of $A$
        appears in $u$. Show that if the complexity function of $u$ satisfies
        \[
          \exists n\in\mathbb{N},\quad p_u(n) < n + \operatorname{Card}(A) - 1,
        \]
        then $u$ is ultimately periodic.

  \item Let $k\ge 1$. Throughout this question, $u=(u_n)_{n\in\mathbb{N}}$ is a
        non-recurrent sequence whose complexity function satisfies
        \[
          p_u(n)=n+k,\qquad \text{for all } n.
        \]
        \begin{enumerate}[label=(\alph*)]
          \item Let $a=u_0$ be the first letter of $u$. The goal is to prove by contradiction
                that the letter $a$ does not extend to the left in $u$.
                Assume, on the contrary, that $a$ extends to the left in $u$.
                What are the two possible shapes of the order-$1$ word graph of $u$?
                Deduce that the $k+1$ letters of $A$ appear infinitely often in $u$.
                Deduce from the non-recurrence of $u$ and Question~2 a contradiction.

          \item What is the complexity of the sequence $(u_n)_{n\ge 1}$?

          \item Deduce that if $u$ is non-recurrent, then there exist $l\ge 1$, a subalphabet $A_0$
                of $A$, letters $a_1,a_2,\dots,a_l \in A$ such that
                $A=A_0\cup\{a_1,\dots,a_l\}$ and a recurrent sequence $w$ over the alphabet $A_0$ such that
                \[
                  u=a_1a_2\cdots a_l w.
                \]
                What is the complexity of $w$?
        \end{enumerate}
\end{enumerate}

\section{Central words}

\begin{enumerate}[label=\arabic*.]
  \item Give the Christoffel word of slope $5/4$ over the alphabet $\{x<y\}$.
        Give its Cayley graph. Give its dual Christoffel word.

  \item Give the inverses of $5$ and $4$ modulo $9$.
        Construct all non-constant words of periods $5$ and $4$ of length $7$.

  \item Show that if two words $U$ and $V$ satisfy $UV=VU$, then there exist a word $W$
        and $i,j\in\mathbb{N}$ such that $U=W^i$ and $V=W^j$.
        Give an inductive proof and a proof based on the Fine--Wilf theorem.
\end{enumerate}

\section{Sturmian words}

Show that the set of Sturmian factors is closed under reversal (mirror image).

\section{Episturmian words}

An infinite word is called \emph{episturmian} if its language is closed under reversal
and it has a unique right special factor of each length.
Show that frequencies are uniform for an episturmian word.

\section{Chacon word}

The Chacon word $u=\sigma^\infty(0)$ is defined over $\{0,1\}$ by the substitution
$\sigma : 0 \mapsto 0010$, $1 \mapsto 1$.
Show that the Chacon word is uniformly recurrent.
\textit{Hint:} Show that the words $b_n$ defined by
\[
  b_0=1,\qquad b_{n+1}=b_n b_n 1 b_n \quad \text{for all } n
\]
are prefixes of $u$.
Show that frequencies are uniform in the Chacon word.

\section{Frequencies}

Consider the substitution $\sigma : 0 \mapsto 0011$, $1 \mapsto 0101$.
Let $u=\sigma^\infty(0)$.
Show that $(O(u),T)$ is minimal and uniquely ergodic.
Do factor frequencies exist?
Show that the values of frequencies on cylinder sets are not determined
by their values on letter-cylinder sets.

\end{document}
